---
title: "Custom fitting with apply_fit()"
author: "gDR team"
output: BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{Custom fitting with apply_fit()}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r knitr-config, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

```{r setup}
library(gDRcore)
library(gDRtestData)
library(gDRutils)
library(SummarizedExperiment)
library(BumpyMatrix)
library(data.table)
```

# Overview

The standard gDR pipeline fits dose-response curves using a fixed
4-parameter log-logistic model (`fit_SE()`).  Sometimes you need something
different: an alternative curve model, custom synergy metrics, Bayesian
estimates, or bespoke pharmacology metrics for your assay.

`apply_fit()` and `apply_fits()` let you plug **any R function**
into the gDR SE/MAE pipeline without touching the pipeline internals.

Key properties:

- Works with **single-agent**, **combination**, and **time-course** data.
- Each fit function writes to its **own named assay** — no risk of overwriting
  native gDR assays (`"Metrics"`, `"scores"`, `"excess"`, …).
- Results are **idempotent**: calling twice with the same `fit_source`
  overwrites rather than duplicates.
- `apply_fits()` applies N fit functions in a **single** BumpyMatrix
  traversal — efficient when you have several metrics to compute on the same
  data.

---

# Data setup

We use a small synthetic single-agent SE from `gDRtestData`.

```{r load-data}
mae <- gDRutils::get_synthetic_data("finalMAE_small.qs2")
sa_se <- mae[["single-agent"]]
sa_se
```

The `Averaged` assay is the input to custom fit functions.  Each cell of the
BumpyMatrix contains one data.table per (drug × cell line) pair:

```{r inspect-averaged}
avg_cell <- BumpyMatrix::unsplitAsDataFrame(
  assay(sa_se, "Averaged"),
  row.field = "row", column.field = "column"
)
head(avg_cell[avg_cell$row == avg_cell$row[1] &
              avg_cell$column == avg_cell$column[1], ])
```

Each cell has four columns:

| Column | Description |
|---|---|
| `normalization_type` | `"GR"` or `"RV"` |
| `Concentration` | drug concentration (µM) |
| `x` | averaged normalized response (GR value or relative viability) |
| `x_std` | standard deviation across replicates |

---

# The fit_fn contract

A fit function must satisfy this interface:

```
fit_fn(avg_dt) -> named list (or single-row data.frame / data.table)
```

`avg_dt` is a **data.table** containing the rows for **one** combination of
slicing column values (by default: one `normalization_type`) for one
(drug × cell line) cell.

The returned named list becomes one row in the output assay.  Column names
are the names of the list elements.  `fit_source` is stamped automatically
by the generic layer — do not include it in the return value.

## Minimal example

```{r minimal-fn}
# The simplest possible fit_fn: compute mean and SD of the response
summary_fn <- function(avg_dt) {
  list(
    x_mean = mean(avg_dt$x, na.rm = TRUE),
    x_sd = sd(avg_dt$x, na.rm = TRUE),
    n = NROW(avg_dt)
  )
}
```

---

# Single-agent: apply_fit()

Apply the summary function to every (drug × cell line × normalization_type)
triplet and write results to a custom assay named `"custom_summary"`.

```{r sa-basic}
sa_out <- apply_fit(
  sa_se,
  fit_fn = summary_fn,
  data_type = "single-agent",
  output_assay = "custom_summary",
  fit_source = "demo"
)
assayNames(sa_out)
```

```{r sa-inspect}
summary_df <- BumpyMatrix::unsplitAsDataFrame(
  assay(sa_out, "custom_summary"),
  row.field = "row", column.field = "column"
)
head(summary_df)
```

One row per (drug × cell line × normalization_type) triplet. The native
`"Metrics"` assay is untouched.

## Idempotent merge

Calling again with the same `fit_source` **replaces** those rows rather than
appending:

```{r sa-idempotent}
n_before <- NROW(BumpyMatrix::unsplitAsDataFrame(
  assay(sa_out, "custom_summary"),
  row.field = "row", column.field = "column"
))

# Call again — same fit_source, same data
sa_out2 <- apply_fit(
  sa_out,
  fit_fn = summary_fn,
  data_type = "single-agent",
  output_assay = "custom_summary",
  fit_source = "demo"
)
n_after <- NROW(BumpyMatrix::unsplitAsDataFrame(
  assay(sa_out2, "custom_summary"),
  row.field = "row", column.field = "column"
))

stopifnot(n_before == n_after)   # no duplicate rows
message("Row count before: ", n_before, " — after: ", n_after, " (no change)")
```

## Coexisting fit sources

Different `fit_source` values live **side by side** in the same assay:

```{r sa-coexist}
extra_fn <- function(avg_dt) list(x_max = max(avg_dt$x, na.rm = TRUE))

sa_two <- sa_out |>
  apply_fit(extra_fn, "single-agent",
                   output_assay = "custom_summary",
                   fit_source = "extremes")

sources <- unique(BumpyMatrix::unsplitAsDataFrame(
  assay(sa_two, "custom_summary"),
  row.field = "row", column.field = "column"
)$fit_source)
message("fit_source values in assay: ", paste(sources, collapse = ", "))
```

---

# Reference Hill fit: fit_drug_response_metrics()

A reference implementation that replicates what `fit_SE()` computes internally
(LL.4 Hill model via `drc`) is provided as `fit_drug_response_metrics()`.
Use it as a starting point for custom Hill-like models or to validate that
your fit function produces consistent results.

```{r hill-ref}
hill_out <- apply_fit(
  sa_se,
  fit_fn = fit_drug_response_metrics,
  data_type = "single-agent",
  output_assay = "custom_hill",
  fit_source = "hill_ref"
)

hill_df <- BumpyMatrix::unsplitAsDataFrame(
  assay(hill_out, "custom_hill"),
  row.field = "row", column.field = "column"
)
head(hill_df[, c("row", "column", "normalization_type",
                  "xc50", "h", "r2", "fit_type")])
```

---

# summary_fn: cell-level aggregation

An optional `summary_fn` is called **once per (drug × cell line) cell** on
all rows produced by `fit_fn` for that cell (one row per normalization type).
It is the right place for metrics that aggregate across normalization types —
for example, whether the result is synergistic across both GR and RV.

```{r summary-fn}
# Aggregate: mean xc50 and flag whether both norm types fitted successfully
hill_summary_fn <- function(fit_dt) {
  list(
    mean_xc50 = mean(fit_dt$xc50, na.rm = TRUE),
    mean_r2 = mean(fit_dt$r2, na.rm = TRUE),
    all_converged = all(fit_dt$fit_type == "DRC4pHillFitModel", na.rm = TRUE)
  )
}

hill_with_summary <- apply_fit(
  sa_se,
  fit_fn = fit_drug_response_metrics,
  data_type = "single-agent",
  output_assay = "custom_hill",
  summary_fn = hill_summary_fn,
  summary_assay = "custom_hill_summary",
  fit_source = "hill_ref"
)
assayNames(hill_with_summary)
```

```{r summary-fn-inspect}
sumdf <- BumpyMatrix::unsplitAsDataFrame(
  assay(hill_with_summary, "custom_hill_summary"),
  row.field = "row", column.field = "column"
)
head(sumdf[, c("row", "column", "mean_xc50", "mean_r2", "all_converged")])
```

One row per (drug × cell line) — regardless of how many normalization types
were fitted.

---

# Combination data: Bliss and HSS

`bliss_fit_fn()` and `hss_fit_fn()` are reference implementations for the
two most common combination synergy metrics.

## Building a minimal combination SE

```{r combo-se}
# Build a synthetic 3x3 combination dose grid (0 = no drug = single-agent edge)
combo_dt <- data.table::CJ(
  row = c("DrugA", "DrugB"),
  column = "CellLine1",
  normalization_type = c("GR", "RV"),
  Concentration = c(0, 0.1, 1.0),
  Concentration_2 = c(0, 0.1, 1.0)
)
set.seed(42L)
combo_dt[, x := pmax(0.05,
  1 - 0.3 * Concentration / (Concentration + 0.5) -
      0.2 * Concentration_2 / (Concentration_2 + 0.5) +
      rnorm(.N, 0, 0.03))]

data_cols <- setdiff(names(combo_dt), c("row", "column"))
combo_bumpy <- BumpyMatrix::splitAsBumpyMatrix(
  combo_dt[, data_cols, with = FALSE],
  row = combo_dt$row, col = combo_dt$column
)
combo_se <- SummarizedExperiment(assays = list(Averaged = combo_bumpy))
```

## Bliss independence score

```{r combo-bliss}
bliss_out <- apply_fit(
  combo_se,
  fit_fn = bliss_fit_fn,
  data_type = "combination",
  output_assay = "custom_bliss",
  fit_source = "bliss"
)

bliss_df <- BumpyMatrix::unsplitAsDataFrame(
  assay(bliss_out, "custom_bliss"),
  row.field = "row", column.field = "column"
)
bliss_df[, c("row", "column", "normalization_type",
             "bliss_score", "bliss_excess_mean", "n_combo_points")]
```

`bliss_score > 0` indicates synergy (combination is more potent than Bliss
independence predicts); `bliss_score < 0` indicates antagonism.

## Highest Single Agent score

```{r combo-hss}
hss_out <- apply_fit(
  combo_se,
  fit_fn = hss_fit_fn,
  data_type = "combination",
  output_assay = "custom_hss",
  fit_source = "hss"
)

hss_df <- BumpyMatrix::unsplitAsDataFrame(
  assay(hss_out, "custom_hss"),
  row.field = "row", column.field = "column"
)
hss_df[, c("row", "column", "normalization_type",
           "hss_score", "hss_excess_mean")]
```

---

# Efficient multi-fit: apply_fits()

When several fit functions operate on the **same input assay**, use
`apply_fits()` to traverse each BumpyMatrix cell **once** and apply
all functions in that single pass.

```{r multi-fit}
combo_multi <- apply_fits(
  combo_se,
  fit_fns = list(
    custom_bliss = bliss_fit_fn,
    custom_hss = hss_fit_fn
  ),
  data_type = "combination",
  fit_source = "synergy_panel"
)
assayNames(combo_multi)
```

Both assays are written in one traversal — equivalent to chaining two
`apply_fit()` calls but without the overhead of a second unsplit +
iteration.

## Shared pre-computation pattern

When two metrics share an expensive intermediate (e.g. fitted single-agent
curves), a single fit function can return a **named list of named lists**
to populate multiple assays from one computation:

```{r shared-precompute, eval = FALSE}
# Each top-level name maps to an output assay; inner lists are the rows
bliss_and_hss_combined <- function(dt) {
  # Expensive step done ONCE per cell
  sa1 <- dt[dt$Concentration_2 == 0 & dt$Concentration > 0, ]
  sa2 <- dt[dt$Concentration == 0   & dt$Concentration_2 > 0, ]

  list(
    custom_bliss = list(bliss_score = mean(sa1$x) - mean(sa2$x)), # simplified
    custom_hss = list(hss_score = min(c(sa1$x, sa2$x)))
  )
}

apply_fits(
  combo_se,
  fit_fns = list(custom_bliss = bliss_and_hss_combined,
                    custom_hss = bliss_and_hss_combined),
  data_type = "combination",
  fit_source = "shared"
)
```

---

# Chaining with pipe

The functions are pipe-friendly — each call returns the updated SE:

```{r pipe-chain}
result_se <- combo_se |>
  apply_fit(bliss_fit_fn, "combination",
                   output_assay = "custom_bliss",
                   fit_source = "bliss") |>
  apply_fit(hss_fit_fn, "combination",
                   output_assay = "custom_hss",
                   fit_source = "hss") |>
  apply_fit(
    function(dt) list(n_obs = NROW(dt)),
    "combination",
    output_assay = "combo_diagnostics",
    fit_source = "qc"
  )

assayNames(result_se)
```

---

# Error handling

By default, a failed cell emits a warning and is skipped
(`on_error = "warn"`).  Use `on_error = "stop"` to halt immediately and
propagate the error — useful when debugging a new fit function.

```{r error-warn}
buggy_fn <- function(dt) {
  if (dt$normalization_type[1] == "GR") stop("GR not supported")
  list(x_rv = mean(dt$x, na.rm = TRUE))
}

se_partial <- withCallingHandlers(
  apply_fit(
    sa_se, buggy_fn, "single-agent",
    output_assay = "rv_only",
    fit_source = "rv_fn",
    on_error = "warn"
  ),
  warning = function(w) {
    message("[caught] ", conditionMessage(w))
    invokeRestart("muffleWarning")
  }
)
# Only RV rows are written; GR cells were skipped with a warning
rv_df <- BumpyMatrix::unsplitAsDataFrame(
  assay(se_partial, "rv_only"),
  row.field = "row", column.field = "column"
)
unique(rv_df$normalization_type)
```

---

# Quick reference

## Function signatures

```{r quick-ref, eval = FALSE}
# Single fit → one output assay
apply_fit(
  se,
  fit_fn,
  data_type = "single-agent", # or "combination", "time-course"
  slicing_cols = NULL, # NULL → data_type default
  slicing_values = NULL, # NULL → all unique values found
  input_assay = NULL, # NULL → data_type default ("Averaged")
  output_assay, # REQUIRED — your assay name
  summary_fn = NULL, # optional cell-level aggregator
  summary_assay = NULL,
  merge = "merge", # or "replace"
  on_error = "warn", # or "stop"
  fit_source                         # REQUIRED — upsert key tag
)

# Multiple fits → one BumpyMatrix pass
apply_fits(
  se,
  fit_fns, # named list: name = output assay, value = fit function
  data_type = "single-agent",
  fit_source,
  ...
)
```

## fit_fn contract

| | |
|---|---|
| **Input** | `data.table` — one BumpyMatrix cell, filtered to one `slicing_cols` value |
| **Output** | Named list → one row in `output_assay`; or named list of named lists for multi-assay pattern |
| **fit_source** | Stamped by the generic layer — do **not** include in the return value |
| **slicing columns** | The caller's value (e.g. `normalization_type`) IS available in the data.table — no need to filter again |

## summary_fn contract

| | |
|---|---|
| **Input** | `data.table` — all rows written by `fit_fn` for one (row × column) cell |
| **Output** | Named list → one row in `summary_assay` |
| **When to use** | Aggregated metrics that span normalization types (e.g. "converged in at least one type?") |

## Reference implementations

| Function | Data type | Key output columns |
|---|---|---|
| `fit_drug_response_metrics()` | single-agent | `xc50`, `h`, `r2`, `x_mean`, `x_AOC`, `fit_type` |
| `bliss_fit_fn()` | combination | `bliss_score`, `bliss_excess_mean`, `n_combo_points` |
| `hss_fit_fn()` | combination | `hss_score`, `hss_excess_mean`, `n_combo_points` |

---

# SessionInfo {-}

```{r session-info}
sessionInfo()
```
